Fractional land cover classification method assessment using PROBA-V satellite data

### 2019-07-29 article/article.tex
Changes from papeeria

% Minimalistic template for a generic article, geared towards MDPI guidelines
% Use PDFLaTeX
\documentclass[a4paper,10pt]{article}
% Packages provided by the MDPI template already
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{lineno}
\usepackage{microtype}
\usepackage{amsmath}
\usepackage{indentfirst}
\usepackage{booktabs} % For \toprule, etc.

@@

\abstract{
The current common practice for producing global land cover maps is to represent land cover classes as discrete units, e.g. a pixel may be labelled as grassland, tree cover, or open water. However, in reality, the area covered by the pixel is hardly ever homogeneous, it may be e.g. trees surrounded by grass next to water. This mismatch between the reality and its representation in maps causes a loss in precision, as only the dominant land cover class can be represented in such ``hard'' classification maps. This problem is worse for coarse resolution land cover maps. A common workaround to this issue is to define mosaic classes, i.e. land cover classes that are combinations of other land cover classes, such as open forest (combination of grassland and tree cover). However, this approach leads to increasingly complex map legends and confusion about the definitions.

@@

\section{Introduction}

\begin{enumerate}
 \item Relevance of fractional LC mapping (link to user requirements and SDGs)
 \item Studies that have looked into fractional LC mapping so far
 \item Available fractional methods that were introduced in literature (wavelet transform etc.)
 \item Methods that are tested in this paper
\end{enumerate}

Global land cover maps

@@

>>>>

% Minimalistic template for a generic article, geared towards MDPI guidelines
% Use PDFLaTeX
\documentclass[a4paper,10pt]{article}
% Packages provided by the MDPI template already
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{lineno}
\usepackage{microtype}
\usepackage{amsmath}
\usepackage{indentfirst}
\usepackage{booktabs} % For \toprule, etc.
\usepackage[sort&compress,sectionbib]{natbib}

@@

\abstract{
\comment[id=dm]{The abstract is still a placeholder}The current common practice for producing global land cover maps is to represent land cover classes as discrete units, e.g. a pixel may be labelled as grassland, tree cover, or open water. However, in reality, the area covered by the pixel is hardly ever homogeneous, it may be e.g. trees surrounded by grass next to water. This mismatch between the reality and its representation in maps causes a loss in precision, as only the dominant land cover class can be represented in such ``hard'' classification maps. This problem is worse for coarse resolution land cover maps. A common workaround to this issue is to define mosaic classes, i.e. land cover classes that are combinations of other land cover classes, such as open forest (combination of grassland and tree cover). However, this approach leads to increasingly complex map legends and confusion about the definitions.

@@

\section{Introduction}
% * <Dainius Masiliūnas> 14:50:56 29 Jul 2019 UTC+0200:
% ~1250 words, 8 paragraphs + objectives

\begin{enumerate}
 \item Relevance of fractional LC mapping (link to user requirements and SDGs)
 \item Studies that have looked into fractional LC mapping so far
 \item Available fractional methods that were introduced in literature (wavelet transform etc.)
 \item Methods that are tested in this paper
 \item Global land cover maps
\end{enumerate}

Global land cover mapping is very important from an ecological and earth systems point of view. Land cover maps are used for a variety of applications, from estimating area covered by forests \citep{bartalev2014probavboreal} to air quality modelling \citep{wiedinmyer2006airquality}. In addition, land cover maps, especially those of fine spatial resolution, have the potential to be used for land cover change monitoring \citep{defourny2012cci}. This is important for a number of land cover classes. For instance, forests are an important terrestrial carbon sink in the global carbon cycle, with different forests having different carbon sink capacities \citep{pan2011large}. Knowing their distribution and changes over time allows for more accurate climate change modelling. Another land cover class, wetlands, is key for maintaining biodiversity due to the uniqueness of wetland ecosystems that allows rare and protected species to live there. Wetlands are protected globally by the Ramsar convention \citep{davis1994ramsar}.

The creation of a global land cover map from remote sensing data has been a goal of many different studies \citep{hansen2000hardtree}. Their results are nowadays freely available as satellite imagery products, such as Global Land Cover 2000 \citep{bartholome2005glc2000}, MODIS land cover \citep{friedl2010modis} and GlobCover \citep{arino2007globcover}. There are also maps based on the fusion of multiple other maps, like Geo-Wiki hybrid \citep{see2015hybrid}, and multiple sensors, like Land Cover CCI \citep{lccciguide} and GlobeLand30 \citep{chen2015globeland30}. However, these land cover maps all have certain drawbacks.

The first common drawback of current global land cover classification maps is that their spatial resolution is low to medium. The majority of aforementioned land cover maps are derived from MODIS and MERIS sensor data. These sensors are only capable of capturing images with the finest pixels representing 300 by 300 metre areas (at the equator). Recent advances in satellite sensors and computing power would allow for land cover classification at a higher spatial and temporal resolution. For example, at the moment the PROBA-V mission by the European Space Agency produces imagery with as fine as 100 by 100 metre pixels (with additional products also providing 300 by 300 metre and 1 by 1 kilometre pixel size for comparison with previous sensors) \citep{probavguide}. It is well-suited for time series analysis, because it has an archive that goes back to 2013, as well as a fast revisit time of 2 days for full global coverage (1 day for locations above 35\textdegree{} latitude) \citep{dierckx2014probav}.

The second drawback of current global land cover products is that their classification accuracy tends to be low, averaging at around 65\% \citep{tsendbazar2016integrating}. The third drawback is that all of these products use what is known as ``hard'' or ``crisp'' classification: each pixel is assigned to one particular land cover class only. The result of a hard classification is a thematic map. This classification type treats each pixel as homogeneous with regards to land cover, which is rarely the case in reality. In contrast, a ``fuzzy'' or ``soft'' classification, sometimes also called ``subpixel'' classification or ``linear mixture modelling'' \citep{Okeke2006fuzzyexponent} identifies the proportion of each land cover class within each pixel. The result of a fuzzy classification is one map per land cover class, showing the proportion of the pixel area covered by the land cover class.

Hard classification is not well-suited for coarse resolution imagery, due to a large proportion of mixed pixels in it (representing areas of mixed land cover), compared to endmember pixels (or pure pixels, which are dominated by a single land cover class). When hard classification is attempted on such imagery, the accuracy of the result can be no higher than the area fraction that the dominant land cover type of the pixel occupies \citep{latifovic2004accuracy}. There have been attempts at increasing accuracy by defining mosaic classes, such as mixed forests (50\% needleleaf and 50\% broadleaf trees), but it leads to a proliferation of classes that increases map complexity and still does not deal with the core problem, thus not improving the classification accuracy by much \citep{tsendbazar2016comparative}. In contrast, ``fuzzy'' or ``soft'' classification results in each pixel containing information about the proportion of each class within that pixel, therefore operating on sub-pixel scales. This has the advantage of representing land cover more accurately, and gives the ability to represent mosaic classes as a combination of pure classes instead. Since the output of fuzzy classification is one raster per land cover class, with smooth transitions between classes, it is suitable for more in-depth analysis and user-specific visualisation criteria \citep{tsendbazar2016integrating}. Despite that, hard classification is still the most often used classification type due to the number of algorithms developed for it, ease of storing the result (it is thematic and thus takes little storage space), displaying it in a single map (albeit in a less accurate fashion) and performing accuracy assessment.

There is a number of different classification algorithms, but only several of them are suitable to be used for fuzzy classification \citep{nath2014methods}. The two methods most commonly used in scientific literature are fuzzy \textit{c}-means and neural networks \citep{zhang2001fullyfuzzy}. Neural networks in particular are well-suited for fuzzy classification, since they allow multiple continuous output as well as input variables in a single model \citep{foody1997fuzzynnet}. Fuzzy \textit{c}-means, also known as fuzzy \textit{k}-means or soft \textit{k}-means, is a statistical method that relies on the proximity of pixels in feature space to class centroids, and thus is also suitable for fully fuzzy classification. However, since \textit{k}-means is an unsupervised classification algorithm, the ability to make use of training data in fuzzy \textit{c}-means is limited to determining class centroids with more precision \citep{hengl2004fuzzycmeans}.

In addition, any other algorithms that provide a measure of uncertainty about class membership can be used, such as the ratio of individual tree votes of random forest \citep{breiman2001random} or class probabilities in gradient boosting \citep{friedman2001gradientboost}. While classification uncertainty is not a direct measure of class membership \citep{sytze2000fuzzyset}, they are nonetheless correlated. As such, it has been used by numerous authors as a proxy indicator of class membership, achieving satisfactory classification accuracy \citep{foody2002accuracy}.

Furthermore, algorithms that can handle continuous variables, but only one response variable (such as random forest), can be used as well, by creating separate models for every class. This approach is called binary relevance \citep{karalas2016br}. Using this approach, the data needs to be post-processed after classification at a pixel level to conform to physical constraints (class membership must be between 0 and 100\% and sum up to 100\%). Random forest has been reported to give higher or equal accuracy results compared to other algorithms, like Support Vector Machines and individual decision trees, in hard classification scenarios using satellite imagery similar to that of PROBA-V \citep{duro2012algorithmcomparison}. Random forest regression was also shown to perform as well as other algorithms in fuzzy classification scenarios \citep{walton2008subpixelrf}, although it is used in much fewer studies on fuzzy classification than the other algorithms mentioned previously. Gradient boosting is an algorithm related to random forest, with the same advantages and disadvantages, but it is known to perform better in machine learning challenges \citep{chen2015higgs} and has not yet been tested on fuzzy land cover classification.

Accuracy assessment is rather straight-forward for hard classification: typically a confusion matrix is employed for this purpose, showing how many pixels (or in some cases how much area \citep{stehman2009sampling}) in the image have been classified correctly, and how many incorrectly. Such a matrix makes it simple to tell which classes are hard to discern from one another, as well as allows for deriving statistics such as users' accuracy, producers' accuracy, and overall accuracy \citep{foody1996fuzzyevaluation}, as well as variance if probability sampling was used. However, a standard confusion matrix is not applicable in the context of fuzzy classification, since misclassification in this case is not absolute, but rather a matter of degree \citep{foody2002accuracy}. Several solutions, such as distance \citep{foody1996fuzzyevaluation}, cross-entropy and mutual information \citep{lu2007methods} have been suggested as alternatives for fuzzy classification accuracy assessment.

Visualisation of the fuzzy classification results is also challenging, since each class effectively is a single-channel raster of its own. Three such classes can easily be combined into RGB channels for visualisation, but with more classes it is no longer possible. There have been attempts to develop a method based on the hue, saturation and intensity colour model to allow for a larger number of distinct classes to be visualised on a single raster \citep{hengl2004fuzzycmeans}. There is also a possibility of ``hardening'' the classification when high accuracy visualisation is not needed, and making use of multiple RGB rasters side-by-side when it is needed.

All in all, even though global land cover mapping has been a focus of many other studies, existing global land cover maps still have room for improvement. This can be achieved by using data from the sensors of newer satellites such as PROBA-V, as well as performing fuzzy classification as opposed to the traditional hard classification.

An important long-term goal of land cover mapping is to have a highly accurate fuzzy land cover classification method at a global scale. Since there are many different methods that could be used to achieve such a goal, and since the amount of data generated by satellites covering the entire world is enormous, this thesis focuses on comparing several different fuzzy classification methods on a subset of global imagery provided by the PROBA-V satellite.

PROBA-V is a relatively new satellite, with few studies using it for land cover classification so far. A number of studies have used simulated PROBA-V data in preparation for its launch \citep{stathakis2014probavurban,roumenina2013probavcrops,bartalev2014probavboreal}. After launch, most studies have focused on its potential for crop classification \citep{roumenina2015probavcrops,durgun2016crop,lambert2016cropland}. All of these studies have only used the traditional hard classification. As such, the potential of using PROBA-V data for fuzzy classification in order to obtain a more accurate representation of land cover at global scale has never been tested before.

Another problem of land cover classification is that there is high disagreement between different land cover products on mixed classes, such as mixed forests \citep{Herold2008lccomparison}. Mixed forests and wetlands are challenging to detect and classify using hard classification and optical remote sensing data, so it is important to check how well large-scale fuzzy classification performs on these classes. Therefore this thesis focused specifically on the region of ecotone from boreal forests to temperate broadleaf forests in Europe. This region also includes a large number of boreal wetlands.

In addition, there is a wide variety of methods that can be used for fuzzy classification. Most studies on this topic focus on comparing only two methods at once. In this thesis, four of them have been tested: fuzzy \textit{c}-means, neural networks, random forest regression and multiclass gradient boosting. All of these methods are based on fundamentally different approaches to classification, so the results give insight into the possible performance of similar methods as well. In addition, the last of the mentioned methods has not yet been tested on fuzzy classification itself (although related methods have been).

When it comes to global image processing, one problem is the amount of data and the time it takes to process it. Some algorithms are known to suffer from the curse of dimensionality, where a linear increase in input data results in a geometric increase in processing time \citep{walton2008subpixelrf}. Therefore it is important to also measure the processing time of different classification methods.

In addition to classification methods, one aspect that is important for classification accuracy is the data (covariates) that the algorithm is trained on and predict land cover from \citep{yu2014metadiscoveries}. In addition to spectral covariates (blue, red, near infrared, shortwave infrared bands), temporal covariates (growing season start, duration, growing intensity) are commonly used to improve the separability of crops \citep{jakubauskas2001harmonic}. Elevation covariates (elevation, slope, aspect) are known to improve the separability between tree classes \citep{burrough2001fuzzy}. Vegetation indices have been used to improve the detection of wetlands \citep{sader1995wetlands}, as well as for defining classification rules in rule-based fuzzy classification \citep{baraldi2006rulebased}. Therefore in this thesis such covariates have also been used in order to increase land cover classification accuracy, and their importance for prediction was analysed.

All in all the goal of this thesis was to improve large-scale land cover maps by performing fuzzy land cover classification in the boreal forest-temperate broadleaf forest gradient zone using PROBA-V satellite imagery. The research questions of the thesis are:

@@

\bibliography{article-bib}
